---
---
title: "Cost-Effectiveness of SGLT2 Inhibitors vs. Placebo for Heart Failure Prevention in Type 2 Diabetes with Established ASCVD"
subtitle: "A Markov model with deterministic and probabilistic sensitivity analysis"
author: "Said Adjao"
date: today
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(tidyverse)

# ---- Settings ----
n_cycles <- 20
discount <- 0.03
states   <- c("Stable", "PostHF", "Dead")
disc     <- 1 / (1 + discount)^(0:n_cycles)
hr       <- 0.70   # from meta-analysis (95% CI 0.62-0.78)

# ---- Base-case parameters ----
p_stable_event_ctrl <- 0.040
p_stable_death      <- 0.020
p_posthf_death      <- 0.060
c_drug   <- 6000; c_event <- 13000; c_stable <- 2000; c_posthf <- 4000
u_stable <- 0.85; u_posthf <- 0.70; u_dead <- 0

# ---- Model functions ----
make_tm <- function(p_event, p_sd, p_pd) {
  matrix(c(1 - p_event - p_sd, p_event, p_sd,
           0, 1 - p_pd, p_pd,
           0, 0, 1),
         nrow = 3, byrow = TRUE, dimnames = list(from = states, to = states))
}
run_markov <- function(tm, n_cycles) {
  tr <- matrix(0, n_cycles + 1, 3, dimnames = list(cycle = 0:n_cycles, state = states))
  tr[1, ] <- c(1000, 0, 0)
  for (t in 1:n_cycles) tr[t + 1, ] <- tr[t, ] %*% tm
  tr
}
tally <- function(trace, cost_vec, qaly_vec) {
  cc <- (trace %*% cost_vec) / 1000
  cq <- (trace %*% qaly_vec) / 1000
  c(cost = sum(cc * disc), qaly = sum(cq * disc))
}

# ---- Base case ----
tm_ctrl <- make_tm(p_stable_event_ctrl,      p_stable_death, p_posthf_death)
tm_trt  <- make_tm(p_stable_event_ctrl * hr, p_stable_death, p_posthf_death)
qaly_vec  <- c(u_stable, u_posthf, u_dead)
cost_ctrl <- c(c_stable, c_posthf, 0)
cost_trt  <- c(c_stable + c_drug, c_posthf + c_drug, 0)
res_ctrl <- tally(run_markov(tm_ctrl, n_cycles), cost_ctrl, qaly_vec)
res_trt  <- tally(run_markov(tm_trt,  n_cycles), cost_trt,  qaly_vec)
d_cost_base <- res_trt["cost"] - res_ctrl["cost"]
d_qaly_base <- res_trt["qaly"] - res_ctrl["qaly"]
icer_base   <- d_cost_base / d_qaly_base
```

## Summary

This analysis evaluates the cost-effectiveness of SGLT2 inhibitors versus
placebo for the prevention of heart failure hospitalization in adults with type
2 diabetes and established atherosclerotic cardiovascular disease (ASCVD), using
a Markov state-transition model. The treatment effect — a hazard ratio of 0.70
for HF hospitalization — was drawn from an independent meta-analysis of four
cardiovascular outcome trials.

At a list price of \$`r format(c_drug, big.mark=",")`/year, the base-case
incremental cost-effectiveness ratio (ICER) was
**\$`r format(round(icer_base), big.mark=",")` per QALY**, exceeding the US
willingness-to-pay threshold of \$150,000/QALY. The conclusion was robust to
uncertainty in the treatment effect but highly sensitive to drug price.

## Background

Heart failure is a frequent, costly complication of type 2 diabetes. SGLT2
inhibitors reduce HF hospitalization, but their value depends on whether the
clinical benefit justifies the cost — the central question for payers and health
technology assessment. This analysis links a meta-analytic estimate of
effectiveness to an economic model to estimate cost per QALY in a defined
high-risk population.

## Methods

### Model structure

A three-state Markov model (Stable; Post-HF hospitalization; Dead) was run over
`r n_cycles` annual cycles with 3% discounting of costs and QALYs. Patients begin
Stable and may transition to Post-HF (a costed event), to Dead (absorbing), or
remain. The treatment effect acts solely by reducing the probability of HF
hospitalization (baseline rate × HR 0.70); all other parameters are identical
between arms.

### Parameters

```{r}
#| label: param-table

tibble(
  Parameter = c("HF hospitalization rate (placebo, annual)",
                "Hazard ratio (SGLT2i vs placebo)",
                "Background mortality (Stable)",
                "Mortality (Post-HF)",
                "Drug cost (annual)", "HF hospitalization cost",
                "Utility: Stable", "Utility: Post-HF",
                "Discount rate", "Time horizon"),
  Value = c("0.040", "0.70 (95% CI 0.62–0.78)", "0.020", "0.060",
            "$6,000", "$13,000", "0.85", "0.70", "3%", "20 years"),
  Source = c("Trial placebo arms", "Meta-analysis (4 RCTs)", "Life tables",
             "Post-HF literature", "Drug pricing (list)", "Costing study",
             "Utility study", "Utility study", "Convention", "—")
) %>% knitr::kable(caption = "Base-case model inputs (illustrative values).")
```

Effectiveness was pooled on the log scale (random-effects); costs and utilities
were sourced from standard references. Note that values here are illustrative;
a definitive analysis would cite a specific source for each.

## Results

### Base case

```{r}
#| label: base-results

tibble(
  Arm = c("Placebo", "SGLT2i", "Incremental"),
  Cost = c(res_ctrl["cost"], res_trt["cost"], d_cost_base),
  QALYs = c(res_ctrl["qaly"], res_trt["qaly"], d_qaly_base)
) %>%
  mutate(Cost = paste0("$", format(round(Cost), big.mark = ",")),
         QALYs = round(QALYs, 3)) %>%
  knitr::kable(caption = "Base-case costs and QALYs per patient (discounted).")
```

The base-case ICER was
**\$`r format(round(icer_base), big.mark=",")` per QALY** — above the US
threshold, indicating SGLT2i is not cost-effective at list price under these
assumptions.

### One-way sensitivity: treatment effect

```{r}
#| label: owsa-hr

run_icer_hr <- function(hr_v) {
  tm_t <- make_tm(p_stable_event_ctrl * hr_v, p_stable_death, p_posthf_death)
  rc <- tally(run_markov(tm_ctrl, n_cycles), cost_ctrl, qaly_vec)
  rt <- tally(run_markov(tm_t,    n_cycles), cost_trt,  qaly_vec)
  (rt["cost"] - rc["cost"]) / (rt["qaly"] - rc["qaly"])
}
tibble(
  Scenario = c("Optimistic (HR 0.62)", "Base case (HR 0.70)", "Pessimistic (HR 0.78)"),
  ICER = sapply(c(0.62, 0.70, 0.78), run_icer_hr)
) %>%
  mutate(ICER = paste0("$", format(round(ICER), big.mark = ","))) %>%
  knitr::kable(caption = "ICER across the 95% CI of the treatment effect.")
```

Across the full confidence interval of the treatment effect, the ICER remained
above \$150,000/QALY. The conclusion is therefore robust to uncertainty in
effectiveness — the result is driven by other parameters, principally drug cost.

### One-way sensitivity: drug price (threshold analysis)

```{r}
#| label: owsa-price
#| fig-width: 8
#| fig-height: 4.5

run_icer_price <- function(price) {
  ct <- c(c_stable + price, c_posthf + price, 0)
  rc <- tally(run_markov(tm_ctrl, n_cycles), cost_ctrl, qaly_vec)
  rt <- tally(run_markov(tm_trt,  n_cycles), ct,        qaly_vec)
  (rt["cost"] - rc["cost"]) / (rt["qaly"] - rc["qaly"])
}
prices <- seq(500, 6000, by = 250)
price_df <- tibble(price = prices, icer = sapply(prices, run_icer_price))
threshold_price <- approx(price_df$icer, price_df$price, xout = 150000)$y

ggplot(price_df, aes(price, icer)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 150000, linetype = "dashed", color = "red") +
  geom_vline(xintercept = threshold_price, linetype = "dotted") +
  scale_x_continuous(labels = scales::dollar) +
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "ICER vs. Annual Drug Price",
       x = "Annual drug cost", y = "ICER ($/QALY)") +
  theme_minimal()
```

SGLT2i reaches the \$150,000/QALY threshold at an annual price of approximately
**\$`r format(round(threshold_price), big.mark=",")`**. Because real-world net
prices (after rebates) often fall near this range, list-price and net-price
analyses can reach opposite conclusions — explaining the divergence between this
model and favorable real-world evaluations.

### Probabilistic sensitivity analysis

```{r}
#| label: psa
#| fig-width: 8
#| fig-height: 4.5

set.seed(123)
n_sims <- 5000
beta_params  <- function(m, n) list(a = m*n, b = (1-m)*n)
gamma_params <- function(m, s) list(shape = (m/s)^2, rate = m/(s^2))

log_hr_se <- (log(0.78) - log(0.62)) / (2 * 1.96)
hr_d   <- exp(rnorm(n_sims, log(0.70), log_hr_se))
pe_d   <- with(beta_params(0.040, 200), rbeta(n_sims, a, b))
sd_d   <- with(beta_params(0.020, 200), rbeta(n_sims, a, b))
pd_d   <- with(beta_params(0.060, 200), rbeta(n_sims, a, b))
us_d   <- with(beta_params(0.85, 100),  rbeta(n_sims, a, b))
up_d   <- with(beta_params(0.70, 100),  rbeta(n_sims, a, b))
cdr_d  <- with(gamma_params(6000, 1200), rgamma(n_sims, shape, rate))
cs_d   <- with(gamma_params(2000, 400),  rgamma(n_sims, shape, rate))
cp_d   <- with(gamma_params(4000, 800),  rgamma(n_sims, shape, rate))

psa <- tibble(d_cost = numeric(n_sims), d_qaly = numeric(n_sims))
for (i in 1:n_sims) {
  tm_c <- make_tm(pe_d[i],          sd_d[i], pd_d[i])
  tm_t <- make_tm(pe_d[i]*hr_d[i],  sd_d[i], pd_d[i])
  qv <- c(us_d[i], up_d[i], 0)
  cc <- c(cs_d[i], cp_d[i], 0)
  ct <- c(cs_d[i] + cdr_d[i], cp_d[i] + cdr_d[i], 0)
  rc <- tally(run_markov(tm_c, n_cycles), cc, qv)
  rt <- tally(run_markov(tm_t, n_cycles), ct, qv)
  psa$d_cost[i] <- rt["cost"] - rc["cost"]
  psa$d_qaly[i] <- rt["qaly"] - rc["qaly"]
}

ggplot(psa, aes(d_qaly, d_cost)) +
  geom_point(alpha = 0.12, color = "steelblue") +
  geom_abline(slope = 150000, intercept = 0, linetype = "dashed", color = "red") +
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "Cost-Effectiveness Plane (5,000 simulations)",
       subtitle = "Dashed line = $150,000/QALY threshold; points below are cost-effective",
       x = "Incremental QALYs", y = "Incremental Cost") +
  theme_minimal()
```

```{r}
#| label: ceac
#| fig-width: 8
#| fig-height: 4.5

thr <- seq(0, 500000, by = 10000)
ceac <- tibble(wtp = thr,
               p = sapply(thr, function(w) mean(psa$d_cost < w * psa$d_qaly)))
p_150 <- round(100 * mean(psa$d_cost < 150000 * psa$d_qaly), 1)

ggplot(ceac, aes(wtp, p)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 150000, linetype = "dashed", color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(title = "Cost-Effectiveness Acceptability Curve",
       x = "Willingness-to-pay ($/QALY)", y = "Probability cost-effective") +
  theme_minimal()
```

Accounting for all parameter uncertainty simultaneously, SGLT2i had an
approximately **`r p_150`% probability** of being cost-effective at
\$150,000/QALY at list price. The acceptability curve crosses 50% near the
base-case ICER, confirming internal consistency.

## Limitations

This is a simplified three-state model with illustrative inputs; a definitive
analysis would source and cite each parameter, model recurrent-event costs
explicitly, and extend the survival structure. The base-case verdict is highly
sensitive to drug price, and list-price assumptions differ materially from
negotiated net prices. Results should be interpreted as a demonstration of
method rather than a definitive coverage recommendation.

## Conclusion

SGLT2 inhibitors substantially reduce HF hospitalization (HR 0.70), but at list
price the incremental cost per QALY exceeds US willingness-to-pay thresholds in
this population. The conclusion is robust to uncertainty in the treatment effect
and driven primarily by drug cost; the drug becomes cost-effective at an annual
price near \$`r format(round(threshold_price), big.mark=",")`. This illustrates a
core HEOR principle: clinical effectiveness does not guarantee cost-effectiveness.